---
title: "Untitled"
output: html_document
---

This file takes raw IR data and caclulates average heart rate every 60 seconds. For periods where the sample period for all 8 sensors is greater than 99ms, that sample is ignored. Currently, when a sample period of greater than 99ms is found, that entire minute of data is ignored and an NA is written to the heart rate file. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
library(pracma) #for findpeaks
library(lubridate) #for working with date/time
fname = file.choose() # store the filename for use
musselheart<-read.csv(fname) #tested with 20200910_0000_00_SN04_IR
# extract serial number from file name
bname = basename(fname)
SN = substr(bname,start = 18, stop = 21) # SNxx is in positions 18-21 in the file names
```

```{r sampleRate}
Fs = 10 # Sampling frequency Fs (samples per second)
samplesPerMinute = Fs * 60 # Number of samples in a complete minute (10Hz * 60 sec = 600 samples)
```

Data wrangling
```{r}
#create new column with sample period
musselheart<-mutate(musselheart, sampleperiod = endMillis - startMillis)
#creates separate frame of all the gaps in sampling where a full sample took more than 99ms. Not exactly useful but good to know how many gaps there were in one day and when they occur
musselheartgaps<-subset(musselheart, sampleperiod>99)
#subset of data without the long sampling periods, also not exactly useful 
musselheartfast<-subset(musselheart, sampleperiod<100)
```

Loop to calculate heart rate for a full day on one sensor. The arguments of findpeaks() in row 55 are likely sensor-specific so testing code towards the end of the file will be useful when calculating heart rate of other sensors
```{r}
#Object creation prior to running loop. reset r, t, and heartrates6 between runs

#r takes the first row of DateTime as our initial time value
#r=as.POSIXct(spyvalve$DateTime[1]) # No spyvalve dataset yet?

# typical Time zone options: etc/GMT+9 is alaska standard time, UTC is greenwich mean 
# time, etc/GMT+8 is alaska daylight time (and Pacific standard time)
r=as.POSIXct(musselheart$DateTime[1], tz = 'etc/GMT+9')
#t is the end of whatever minute r is, these are the book ends of the minute of data we'll look at
# LPM: POSIXct values are stored internally as number of seconds since 1970-1-1 (i.e. unixtime)
# so you can just add 59 to r and you'll have a time 59 seconds in the future
#t=r+seconds(59)
t=r+59
#update DateTime from class factor to POSIXct so that we can use relational operators in our loop
# typical Time zone options: etc/GMT+9 is alaska standard time, UTC is greenwich mean 
# time, etc/GMT+8 is alaska daylight time (and Pacific standard time)
musselheart$DateTime = as.POSIXct(musselheart$DateTime, tz = 'etc/GMT+9')
#create column names for the heartrates6 data frame
l <- c('UnixTime','SN04S6_bpm')
#create the heartrates6 data frame. For some reason the column names get overwritten if nrow = 0, so it makes row 1 with NAs
heartrates6 <- as.data.frame(matrix(ncol = length(l), nrow=1, dimnames = list(NULL,l)))

for (i in 1:(nrow(musselheart)/600)) {
  #create a subset of one minute of data 
  musselheart_sub = filter(musselheart, DateTime <= t & DateTime >= r)
  musselheart_sub = musselheart[musselheart$DateTime >= r & musselheart$DateTime<=t,]
  for (j in 1:nrow(musselheart_sub)) {
  #create a vector that saves a 1 for rows where sampleperiod < 99 is true and     saves a 0 when false  
  samplelengths<-as.numeric(musselheart_sub$sampleperiod < 99)
  #sums the sample lengths as a single value, to be used in the if statement
  sum_samplelengths <- sum(samplelengths)
    #if sum_samplelengths is 600, then all of our rows have sampleperiod < 99 and     we can calculate HR for this minute
    if(sum_samplelengths == 600) {
    #create vector of IR only
    s6sub<-musselheart_sub$Sensor6IR
    #find peaks in the newly created vector
    beats1<-as.data.frame(findpeaks(-s6sub, nups = 1, ndowns = 2, minpeakdistance = 17,  sortstr = F)) 
    #sort beats1 in order of indices
    beats1<-beats1[order(beats1$V2),]
    #calculate average bpm for the chunk
    y<-mean(60/(diff(beats1$V2)/10)) 
    #add UnixTime value and the average bpm value to newrow1
    newrow1<-c(musselheart_sub[1,]$UnixTime, y)
    #write newrow1 to the heartrates6 data frame
    heartrates6<-rbind(heartrates6,newrow1)  
    #reset y and increment time variables for next iteration
    y = 0
    r = r+minutes(1) 
    t = t+minutes(1)
     } else {
       #if the sample period at any point within that minute (r to t) is greater        than 99ms,write an NA to the data frame and increment to the next minute 
       newrow<-c(musselheart_sub[1,]$UnixTime, NA)
       heartrates6<-rbind(heartrates6, newrow)
       r = r+minutes(1) 
       t = t+minutes(1)
     }
  break
  }
}
#this loop works, but it doesn't stop executing when i gets to nrow(musselheart)/600 and it doesn't stop at all until terminated. This makes heartrates6 a data frame with all the heart rate data, and a bunch of NA rows following it until the loop is terminated. 
```


```{r alternateLoopVersion}
# Data are in musselheart, complete with gaps

# Declare which sensor column you want to use. The exact location of the sensor's column
# may shift day to day, so it's safest to address it directly by this name rather than
# a column number
sensorName = 'Sensor6IR'

# Convert the values in DateTime column to POSIXct
# Typical Time zone options: etc/GMT+9 is alaska standard time, UTC is greenwich mean 
# time, etc/GMT+8 is alaska daylight time (and Pacific standard time)
musselheart$DateTime = as.POSIXct(musselheart$DateTime, tz = 'etc/GMT+9')

# Get the time value of the first minute of the data set
t1 = musselheart$DateTime[1]
t1 = as.POSIXlt(t1)
t1$sec = 0 # Set the seconds value to zero if it's not already
t1 = as.POSIXct(t1) # convert back to POSIXct

# Generate a vector of 1-minute intervals from the start to end time values for this
# data set
timevals = seq.POSIXt(from = t1, 
                      to = musselheart$DateTime[nrow(musselheart)], by = 'min' )
# Make an output dataframe consisting of the time values (1 minute intervals) and 
# column(s) of NAs for the sensor channel(s)
heartrates = data.frame(DateTime = timevals, Sensor = NA)
# Rename the columns.  NOTE: if time stamps are in Alaska standard time, note that here in the
# DateTime column name. If they're in some other time zone (AK daylight savings for example),
# change the column name to reflect that (or shift the timezone of the timestamps to match
# AK standard time, since it's probably the most intuitive time zone to work in with the enviro data)
# Also paste together a sensor column name using the file's serial number info and the sensor's
# column name
sensorColname = paste0(SN,sensorName)
colnames(heartrates) = c('DateTimeAST', sensorColname)

for (i in 1:nrow(heartrates)){
  # See how many rows fall within the time[i] to time[i]+60 range. Ideally this 
  # should be 10Hz * 60 seconds = 600 = samplesPerMinute variable 
  minuteLength = length(musselheart$DateTime[ (musselheart$DateTime >= heartrates$DateTime[i]) & 
                        (musselheart$DateTime < (heartrates$DateTime[i]+60) )  ])
  if (minuteLength == samplesPerMinute) {
    # Good minute's worth of data, do the peak finding
    #create vector of IR only
    sensor_sub = musselheart[(musselheart$DateTime >= heartrates$DateTime[i]) & 
                    (musselheart$DateTime < (heartrates$DateTime[i]+60) ), 
                    sensorName]
    #find peaks in the newly created vector
    beats1<-as.data.frame(findpeaks(-sensor_sub, nups = 1, ndowns = 2, 
                                    minpeakdistance = 17,  sortstr = F))
    #sort beats1 in order of indices
    beats1<-beats1[order(beats1$V2),]
    #calculate average bpm for the chunk
    bpm<-mean(60/(diff(beats1$V2)/10))
    bpm = round(bpm, 1)  # round off result to 1 decimal place
    # Write result to output data frame
    heartrates[i, sensorColname] = bpm
    
  } else if (minuteLength < samplesPerMinute){
    # Can't use this minute's data, advance to next minute
    # Leave the NA in the output data frame for this row
  }
}  # end of for loop



```


Code chunks that follow are earlier ideas and/or options for testing and verification

Initial plotting
```{r}
#plotting 600 data points = 60 seconds of heart beats. Plot each sensor and see which have clear signals
ggplot()+
  geom_line(data = musselheartfast[1:600,], aes(x = startMillis, y = Sensor6IR),       color = "blue")+
  ggtitle("30 Seconds of  Heart Beats")+
  theme_classic()
```

Detect valleys using findpeaks() and verify accuracy by plotting
```{r}
#create vector of IR only for the desired sensor
s8sub<-musselheartfast$Sensor8IR[1:300]
#find valleys in the newly created vector: the arguments of this function can be adjusted to match the signal of each sensor
beats<-as.data.frame(findpeaks(-s8sub, nups = 1, ndowns = 2, minpeakdistance = 10,  sortstr = F))
#plots IR vector and the index values from beats as points. Tweak findpeaks() arguments if needed
plot(s8sub, type = 'l')
points(beats[, 2], s8sub[beats[, 2]], col = "red")
```

Creating the csv with multiple sensors
```{r}

#combine all sensors from the same day into one data frame, do this by making a separate heartrates df for each sensor and then left join them together by UnixTime
SN03_heartrates<-left_join(SN03_heartrates, heartrates, by = "UnixTime")
#save new dataframe as csv
write.csv(SN03_heartrates,"/Users/gabbyk/Downloads/SN03_heartrates.csv", row.names = FALSE)