---
title: "Untitled"
output: html_document
---

This file takes raw IR data and caclulates average heart rate every 60 seconds. For periods where the sample period for all 8 sensors is greater than 99ms, that sample is ignored. Currently, when a sample period of greater than 99ms is found, that entire minute of data is ignored and an NA is written to the heart rate file. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
library(pracma) #for findpeaks
library(lubridate) #for working with date/time
musselheart<-read.csv(file.choose()) #tested with 20200910_0000_00_SN04_IR
```

Data wrangling
```{r}
#create new column with sample period
musselheart<-mutate(musselheart, sampleperiod = endMillis - startMillis)
#creates separate frame of all the gaps in sampling where a full sample took more than 99ms. Not exactly useful but good to know how many gaps there were in one day and when they occur
musselheartgaps<-subset(musselheart, sampleperiod>99)
#subset of data without the long sampling periods, also not exactly useful 
musselheartfast<-subset(musselheart, sampleperiod<100)
```

Loop to calculate heart rate for a full day on one sensor. The arguments of findpeaks() in row 55 are likely sensor-specific so testing code towards the end of the file will be useful when calculating heart rate of other sensors
```{r}
#Object creation prior to running loop. reset r, t, and heartrates6 between runs

#r takes the first row of DateTime as our initial time value
r=as.POSIXct(spyvalve$DateTime[1])
#t is the end of whatever minute r is, these are the book ends of the minute of data we'll look at
t=r+seconds(59)
#update DateTime from class factor to POSIXct so that we can use relational operators in our loop
musselheart$DateTime = as.POSIXct(musselheart$DateTime)
#create column names for the heartrates6 data frame
l <- c('UnixTime','SN04S6_bpm')
#create the heartrates6 data frame. For some reason the column names get overwritten if nrow = 0, so it makes row 1 with NAs
heartrates6 <- as.data.frame(matrix(ncol = length(l), nrow=1, dimnames = list(NULL,l)))

for (i in 1:(nrow(musselheart)/600)) {
  #create a subset of one minute of data 
  musselheart_sub = filter(musselheart, DateTime <= t & DateTime >= r)
  for (j in 1:nrow(musselheart_sub)) {
  #create a vector that saves a 1 for rows where sampleperiod < 99 is true and     saves a 0 when false  
  samplelengths<-as.numeric(musselheart_sub$sampleperiod < 99)
  #sums the sample lengths as a single value, to be used in the if statement
  sum_samplelengths <- sum(samplelengths)
    #if sum_samplelengths is 600, then all of our rows have sampleperiod < 99 and     we can calculate HR for this minute
    if(sum_samplelengths == 600) {
    #create vector of IR only
    s6sub<-musselheart_sub$Sensor6IR
    #find peaks in the newly created vector
    beats1<-as.data.frame(findpeaks(-s6sub, nups = 1, ndowns = 2, minpeakdistance     = 17,  sortstr = F)) 
    #sort beats1 in order of indices
    beats1<-beats1[order(beats1$V2),]
    #calculate average bpm for the chunk
    y<-mean(60/(diff(beats1$V2)/10)) 
    #add UnixTime value and the average bpm value to newrow1
    newrow1<-c(musselheart_sub[1,]$UnixTime, y)
    #write newrow1 to the heartrates6 data frame
    heartrates6<-rbind(heartrates6,newrow1)  
    #reset y and increment time variables for next iteration
    y = 0
    r = r+minutes(1) 
    t = t+minutes(1)
     } else {
       #if the sample period at any point within that minute (r to t) is greater        than 99ms,write an NA to the data frame and increment to the next minute 
       newrow<-c(musselheart_sub[1,]$UnixTime, NA)
       heartrates6<-rbind(heartrates6, newrow)
       r = r+minutes(1) 
       t = t+minutes(1)
     }
  break
  }
}
#this loop works, but it doesn't stop executing when i gets to nrow(musselheart)/600 and it doesn't stop at all until terminated. This makes heartrates6 a data frame with all the heart rate data, and a bunch of NA rows following it until the loop is terminated. 
```

Code chunks that follow are earlier ideas and/or options for testing and verification

Initial plotting
```{r}
#plotting 600 data points = 60 seconds of heart beats. Plot each sensor and see which have clear signals
ggplot()+
  geom_line(data = musselheartfast[1:600,], aes(x = startMillis, y = Sensor6IR),       color = "blue")+
  ggtitle("30 Seconds of  Heart Beats")+
  theme_classic()
```

Detect valleys using findpeaks() and verify accuracy by plotting
```{r}
#create vector of IR only for the desired sensor
s8sub<-musselheartfast$Sensor8IR[1:300]
#find valleys in the newly created vector: the arguments of this function can be adjusted to match the signal of each sensor
beats<-as.data.frame(findpeaks(-s8sub, nups = 1, ndowns = 2, minpeakdistance = 10,  sortstr = F))
#plots IR vector and the index values from beats as points. Tweak findpeaks() arguments if needed
plot(s8sub, type = 'l')
points(beats[, 2], s8sub[beats[, 2]], col = "red")
```

Creating the csv with multiple sensors
```{r}

#combine all sensors from the same day into one data frame, do this by making a separate heartrates df for each sensor and then left join them together by UnixTime
SN03_heartrates<-left_join(SN03_heartrates, heartrates, by = "UnixTime")
#save new dataframe as csv
write.csv(SN03_heartrates,"/Users/gabbyk/Downloads/SN03_heartrates.csv", row.names = FALSE)